Fusing Filters with Integer Linear Programming

### 2014-04-18 2014betterfusionforfilters/section/01-Introduction.tex
cnf: describe format of data flow graphs
cnf: copy edits, add types for array combinators, say the CNF program is in a host program

However, many programs \emph{do} contain opportunities for fusion, if we only knew which opportunities to take. The following example offers \emph{several} unique, but mutually exclusive approaches to fusion. Figure~\ref{f:normalize2-cluterings} on the next page shows some of the possibilities.
\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (> 0)    xs
        sum2 = fold   (+)  0   gts
        ys1  = map    (/ sum1) xs
        ys2  = map    (/ sum2) xs
    in (ys1, ys2)
\end{code}

In Figure~\ref{f:normalize2-cluterings}, the dotted lines show possible clusterings of operators. Stream fusion implicitly choses the solution on the left as its compilation process cannot fuse a produced array into multiple consumers. The best existing ILP approach will chose the solution on the right as it cannot cluster operators that consume arrays of different sizes. Our system choses the solution in the middle, which is also optimal for this example.

>>>>

However, many programs \emph{do} contain opportunities for fusion, if we only knew which opportunities to take. The following example offers \emph{several} unique, but mutually exclusive approaches to fusion. Figure~\ref{f:normalize2-clusterings} on the next page shows some of the possibilities.
\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (> 0)    xs
        sum2 = fold   (+)  0   gts
        ys1  = map    (/ sum1) xs
        ys2  = map    (/ sum2) xs
    in (ys1, ys2)
\end{code}

In Figure~\ref{f:normalize2-clusterings}, the dotted lines show possible clusterings of operators. Stream fusion implicitly choses the solution on the left as its compilation process cannot fuse a produced array into multiple consumers. The best existing ILP approach will chose the solution on the right as it cannot cluster operators that consume arrays of different sizes. Our system choses the solution in the middle, which is also optimal for this example.

### 2014-04-18 2014betterfusionforfilters/section/02-CNF.tex
cnf: copy edits, add types for array combinators, say the CNF program is in a host program

% -----------------------------------------------------------------------------
\section{Combinator Normal Form}
Each input program is expressed in \emph{Combinator Normal Form} (CNF), which is a textual description of its data flow graph. The grammar for CNF is in Figure~\ref{f:CombinatorNormalForm}. Both the @normalize@ and @normalize2@ examples on the previous page are in CNF, and the matching data flow graph for @normalize2@ is in Figure~\ref{f:normalize2-clusterings}. Our data flow graphs are similar to Loop Communication Graphs (LCGs) from related work in imperative array fusion \CITE. We name edges after the corresponding variable from the CNF form, and edges which are fusion preventing are drawn with a dash through them (as per the edge labeled @sum1@ in Figure~\ref{f:normalize2-clusterings}).

@@

\TODO{blend this in.}
The main restriction is that worker functions may only reference scalar variables, nor may they include any array combinators.

We support the following combinators: $@map@_n$, @filter@, @fold@, @gather@ and @cross@.
Most of these are standard combinators except for @gather@ and @cross@, however $@map@_n$ differs slightly from usual @map@ in that it takes $n$ arrays of the same length, and applies the worker function to all elements of the same index. The @gather@ combinator is equivalent to @gather xs ys = map (index xs) ys@. However, as we support no @index@ operation, and worker functions may not refer to arrays, @gather@ is implemented as a primitive.
The @cross@ combinator takes two arrays, and returns the cartesian product of the two.

Since it is unlikely that an entire function will be comprised of these few combinators, we support one additional binding type: @external@.This signifies that the referenced variables are used by a computation that is not a primitive combinator, and must be materialised fully in memory at this stage. The @external@ also lists the variables produced by the external computation. Without knowing the nature of the computation expressed by @external@, we must naturally take a conservative view, and allow no fusion to occur between at these points. They are, in effect, fusion barriers, forcing arrays and scalars to be fully computed before continuing. This also allows us to support combinators other than those here, but as they must be treated conservatively will likely be less optimal than if they were implemented as primitives.

It is important to note that because of the purity of Haskell, we are free to take certain liberties when fusing the program.
None of the worker functions, nor any @external@ computations may produce visible side-effects; the only observable effect must be to produce their output.
This means that any reordering of the program is allowed, as long as mentioned variables are bound beforehand.
\input{figures/Program.tex}

>>>>

% -----------------------------------------------------------------------------
\section{Combinator Normal Form}
Each input program is expressed in \emph{Combinator Normal Form} (CNF), which is a textual description of its data flow graph. The grammar for CNF is in Figure~\ref{f:CombinatorNormalForm}. Both the @normalize@ and @normalize2@ examples on the previous page are in CNF, and the matching data flow graph for @normalize2@ is in Figure~\ref{f:normalize2-clusterings}. Our data flow graphs are similar to Loop Communication Graphs (LCGs) from related work in imperative array fusion \CITE. We name edges after the corresponding variable from the CNF form, and edges which are fusion preventing are drawn with a dash through them (as per the edge labeled @sum1@ in Figure~\ref{f:normalize2-clusterings}). In data flow graphs we tend to elide the worker functions to combinators when they are not important to the discussion --- so we don't show the @(+)@ operator on each use of @fold@.

@@

Most of our array combinators are standard, and suggestive types are given at the bottom of Figure~\ref{f:CombinatorNormalForm}. The $@map@_n$ combinator takes a worker function, $n$ arrays of the same length, and applies the worker function to all elements at the same index. As such it is similar to Haskell's @zipWith@, except for the length restriction on the argument arrays. The @generate@ combinator takes an array length and a worker function, and creates a new array by applying the worker to each index. The @gather@ combinator takes an array of elements, an array of indices, and produces the array of elements that were at each index. In Haskell this would be @gather arr ixs = map (index arr) ixs@. The @cross@ combinator returns the cartesian product of two arrays.

The exact form of the worker functions is left unspecified, as it is not important for the discussion. We assume workers are pure, can at least compute arithmetic functions of their scalar arguments, and index into arrays in the environment. We assume that each CNF program considered for fusion is embedded in a larger ``host program'' which handles file IO and the like. Workers are additionally restricted so they can only directly reference the \emph{scalar} variables bound by the local CNF program, though they may reference array variables bound by the host program. All access to locally bound array variables is via the formal parameters of array combinators, which ensures that all data dependencies we need to consider for fusion are explicit in the data flow graph.

The @external@ binding invokes a host library function that can produce and consume arrays, but not be fused with other combinators. All arrays passed to, and returned from host functions are full materialised. External bindings are explicit \emph{fusion barriers}, which force arrays and scalars to be fully computed before continuing.

% When the function computed is the identity, they can also be used to force a particular intermediate array to be materialized, which provides the programmer fine-grained control over exactly what clustering is produced by our algorithm.

\input{figures/Program.tex}
